##Figure_2

#Libraries
```{r}
library(here)
library(tidyverse)
library(factoextra)
library(MASS)
library(caret)
library(Boruta)
library(ComplexHeatmap)
library(circlize)
library("JLutils")
library(rstatix)
library("fmsb")
library(lme4)
library(glmmTMB)
library(performance)
library(emmeans)
library(psych)
library(rmcorr)
library(ggcorrplot)
library(caroline)
library(umap)
library(plotly)
```

#Colors
```{r}
patient_colors <- c("Patient 1"="#b15928", "Patient 2" = "#ffff99", "Patient 3" = "#6a3d9a",
  "Patient 4" = "#cab2d6", "Patient 5" = "#ff7f00", "Patient 6" = "#fdbf6f",
  "Patient 7" = "#e31a1c", "Patient 8" = "#fb9a99", "Patient 9" = "#34a02c",
  "Patient 10"="#b2df8a", "Patient 11" = "#1f78b4", "Patient 12" = "#a6cee3")

cluster_colors <- c("1" = "#007BFF", "2"="#DC143C", "3"="#fdd4b3")

patient_group_colors <- c("CTRL" = "#390F6E", "AD" = "#D6390E", "CJD" = "#C6D300")
```


#Data - CellProfiler and GeoMx 
```{r}
maindir <- here("Data", "Cellprofiler (morphology)", "Feature selected profiles")
single_profile <- read.csv(here(maindir, "SingleMicroglia_profile.csv"))
labels <- single_profile[,1:22]
features <- single_profile[23:ncol(single_profile)]

geomx_data <- read.csv(here("Data", "GeoMx (protein)", "3. HK normalized Log2 Transformed GeoMx Data.csv"))
names(geomx_data) <- gsub("\\.", " ", names(geomx_data))
geomx_data <- geomx_data %>% select(-S6, -GAPDH, -`Histone H3`)#Remove Housekeepers
```


###Part 1 - Clustering and Feature Importance Evaluation

##Evaluate Optimal Number of Clusters
```{r}
#Elbow plot
elbow <- fviz_nbclust(features,FUN = hcut, method = "wss", title = "Elbow Plot") + theme(text = element_text(size = 18))
elbow + ggtitle("")
```

```{r}
#Silhouette plot
silhouette <- fviz_nbclust(features, FUN = hcut, method = "silhouette") + theme(text = element_text(size = 20))
silhouette + ggtitle("")
```


#Clustering
```{r}
d <- get_dist(features, method="spearman") # Dissimilarity Matrix - Spearman for non-normal distributed morphometric features
hc <- hclust(d, method = "ward.D2" ) # Clustering with Ward Type 2 linkage

plot(hc) #Visualize dendogram of the single IBA1 Objects
sub_group <- cutree.order(hc, k = 3)  #Cut the clustering into desired number of k

table(sub_group) # Number of objects in each group

#Visualization
rect.hclust(hc, k = 3, border = 2:9)

#Save
single_profile <- single_profile %>% #Transfer the cluster classification back to the original single object dataset
  mutate(Cluster = sub_group, .after=Slide)
single_profile$Cluster <- as.factor(single_profile$Cluster)
```

#UMAP
```{r}
set.seed(42)
umap3D = umap(features, n_components = 3, random_state = 15)
#Prepare plot
layout <- data.frame(umap3D[["layout"]]) 
final <- cbind(layout,single_profile$Cluster)
colnames(final) <- c("UMAP1", "UMAP2", "UMAP3", "Cluster")
#UMAp plot
ggplot(final, aes(x = UMAP2, y = UMAP3, color = Cluster)) +
  geom_point(size = 0.5) +
  scale_color_manual(values = cluster_colors, labels = c("1", "2.1", "2.2")) +
  theme_minimal()+
  guides(color = guide_legend(override.aes = list(size = 2.5)))
```

#Evaluation of Clustering using Jackknife Cross Validation
```{r}
Cross_validation <- lda(x = single_profile[24:ncol(single_profile)], grouping = single_profile$Cluster, CV = TRUE)
confusionMatrix(Cross_validation$class, as.factor(single_profile$Cluster))
```

#BORUTA - Identification of most discriminating features
```{r}
#Prepare data
combine_for_Boruta <- cbind(Cluster = single_profile$Cluster, features)
combine_for_Boruta$Cluster <- as.factor(combine_for_Boruta$Cluster)
#Run Boruta
boruta.train <- Boruta(Cluster~., data = combine_for_Boruta, doTrace = 2)
print(boruta.train)
boruta.bank <- TentativeRoughFix(boruta.train)
print(boruta.bank)

#Visualize
#NB!: for better visualization, run the code in the .R script!
par(mar = c(17,2,2,2))
plot(boruta.bank, xlab = "", xaxt = "n")
lz <- lapply(1:ncol(boruta.bank$ImpHistory),function(i)
  boruta.bank$ImpHistory[is.finite(boruta.bank$ImpHistory[,i]),i])
names(lz) <- colnames(boruta.bank$ImpHistory)
Labels <- sort(sapply(lz,median))
axis(side = 1,las=2,labels = names(Labels),
     at = 1:ncol(boruta.bank$ImpHistory), cex.axis = 0.65)
Labels
```


#Visualization of Clusters and most Discriminating Features
```{r}
heat_data <- single_profile
heat_data <- heat_data %>% 
  mutate(Cluster = recode(Cluster, "1"="1", "2"="2.1", "3"="2.2"))

heat_features <- t(features)

#Prepare data
#Colors
heatmap_color_u <- colorRamp2(breaks = c(2, 0, -2), colors = c("yellow", "black", "blue"))
annotation_colors <- list(
  "IBA1+ Object Cluster" = c("1" = "#007BFF","2.1"="#DC143C", "2.2"="#fdd4b3"))

#Annotations
ha <- HeatmapAnnotation("IBA1+ Object Cluster" = heat_data$Cluster,
                        col = annotation_colors, show_legend = c(FALSE),
                        annotation_name_gp = gpar(fontface="bold"))


#Rows to highlight (Discriminating Features evaluated through Boruta)
rows_to_highlight <- c("Total_Texture_InfoMeas1_IBA1_3_03_256", "Total_AreaShape_MajorAxisLength", "Soma_AreaShape_Zernike_0_0", 
                       "Total_Texture_SumEntropy_IBA1_3_02_256", "Soma_AreaShape_Solidity", "Total_AreaShape_Extent")

# Set stylings for row names and make selected rows unique
row_idx <- which(rownames(heat_features) %in% rows_to_highlight)
fontsizes <- rep(4, nrow(heat_features))
fontsizes[row_idx] <- 5
fontcolors <- rep('black', nrow(heat_features))
fontcolors[row_idx] <- 'red'
fontfaces <- rep('plain',nrow(heat_features))
fontfaces[row_idx] <- 'bold'
rowAnno <- rowAnnotation(rows = anno_text(rownames(heat_features), gp = gpar(fontsize = fontsizes, fontface = fontfaces, col = fontcolors)))

# Reorder Dendrogram
dend <- reorder(as.dendrogram(hclust(get_dist(heat_features, method = "spearman"))), -rowMeans(heat_features), agglo.FUN = mean)

# Find Specified features in dendrogram
dend_idx <- which(order.dendrogram(dend) %in% which(rownames(heat_features) %in% rows_to_highlight))

# Draw the heatmap 
htmp <- Heatmap(heat_features, name="Z-score", 
                column_title = "Single IBA1+ Objects", row_title = "Morphometric Features",
                top_annotation = ha, 
                col = heatmap_color_u,
                cluster_rows = dend, right_annotation = rowAnno, show_row_names = F,
                column_split = 3, 
                clustering_method_columns = "ward.D2",
                clustering_method_rows = "ward.D2",
                clustering_distance_columns = "spearman",
                clustering_distance_rows = "spearman",
                heatmap_legend_param = list(legend_height = unit(60, "cm"),  # Adjust the height of the legend
                                            legend_direction = "horizontal",  # Place the legend horizontally
                                            title_position = "topcenter"))


htmp
```

#Visualisation of medians of the  Discriminating Features for each Cluster - Spider Charts
```{r}
features_of_interest <- c("Total_Texture_InfoMeas1_IBA1_3_03_256", "Total_AreaShape_MajorAxisLength", "Soma_AreaShape_Zernike_0_0", 
                         "Total_Texture_SumEntropy_IBA1_3_02_256", "Soma_AreaShape_Solidity", "Total_AreaShape_Extent") #Based of Borutas Selection

#To long format
morph_profile <- cbind(Cluster = single_profile[,"Cluster"], single_profile[,features_of_interest])
morph_profile <- morph_profile%>%
  pivot_longer(!Cluster, names_to="features", values_to = "value")

#Get median values of features of interest for each cluster
morph_summary <- morph_profile %>%
  group_by(Cluster, features) %>%
  get_summary_stats(value, type="median")
morph_summary <- morph_summary[,-c(3,4)]

#Prepare data for Spider Charts
morph_summary <-spread(morph_summary, "features", "median")
morph_summary <- morph_summary %>% remove_rownames %>% column_to_rownames(var="Cluster")

#Identify maximal and minimal values
max <- rep(1, ncol(morph_summary))
min <- rep(-1, ncol(morph_summary))

max_min <- as.data.frame(rbind(max, min))
rownames(max_min) <- c("Max", "Min")
colnames(max_min) <- features_of_interest
morph_summary <- rbind(max_min, morph_summary)

#Rename Morphometric Features (Same order as previously defined)
colnames(morph_summary) <- c("Total - Texture InfoMeas1(.3)", "Total - Major Axis Length", "Soma - Zernike 0.0", 
                               "Total - Texture Sum Entropy(.2)", "Soma - Solidity", "Total - Extent")


#Spider Charts settings
titles <- c("IBA1+ Object Cluster 1", "IBA1+ Object Cluster 2.1", "IBA1+ Object Cluster 2.2")#CHANGE NAMES

#Visualize and save spider charts
op <- par(mar = c(1, 1, 6, 1))
par(mfrow = c(1,3))
for(i in 1:3){
  radarchart(
    df = morph_summary[c(1, 2, i+2), ], vlabels = colnames(morph_summary), 
    axistype = 1, caxislabels = c(-1, -0.5, 0, 0.5, 1), pcol = cluster_colors[i],
    pfcol = scales::alpha(cluster_colors[i], 0.5), plwd = 1.5, plty = 1.5,
    cglcol = "lightgrey", cglty = 1, cglwd = 0.8, calcex = 1,
    axislabcol = "#1C1C1C", vlcex = 0.6, title = titles[i], cex.main = 1.5
  )
}
par(op)
par(mfrow = c(1,1))
```


#Save new single_profile
```{r}
write.csv(single_profile, file=file.path(maindir, "SingleMicroglia_profile_with_cluster.csv"),
          row.names = FALSE) 
```


## Part II - Distribution Analysis
#Cluster Densities
```{r}
total_cells_per_image <- single_profile %>%
  group_by(ImageNumber, Metadata_Sample_number, Patient, Metadata_Pt_type, Metadata_Cortex_region, Metadata_Layer, FileName_ROI_Mask)%>% 
  count() %>% mutate(FileName_ROI_Mask = substr(FileName_ROI_Mask, 6, nchar(FileName_ROI_Mask) - 4)) #Cluster Object Count
colnames(total_cells_per_image)[8] <- "n_cells_in_image" #rename
colnames(total_cells_per_image)[7] <- "Region" #rename

total_clusters_per_image <- single_profile %>% 
  group_by(ImageNumber, Cluster) %>% 
  count() %>% 
  ungroup() %>% 
  complete(
    ImageNumber, 
    Cluster = factor(1:3),
    fill = list(n = 0)
  ) # Long format of n clusters per image

total_clusters_per_image <- merge(total_cells_per_image, total_clusters_per_image, by = "ImageNumber")#Merge
colnames(total_clusters_per_image)[10] <- "n_clustercells_in_image"#Rename

#Adjust for area
area_numbers <- read.csv(here("Data", "Cellprofiler (morphology)", "ROI Areas.csv"), sep=";")
area_numbers$Area <- area_numbers$Area/1000000 # From um to mm

total_clusters_per_image <- merge(total_clusters_per_image, area_numbers, by = "Region")
total_clusters_per_image$cell_per_area <- (total_clusters_per_image$n_clustercells_in_image/total_clusters_per_image$Area)
total_clusters_per_image$cell_per_area <- round(total_clusters_per_image$cell_per_area) #Round to true count data


#Modelling
statistics_n_cluster <- total_clusters_per_image %>%
  group_by(Cluster) %>%
  group_modify(~ {
    current_cluster <- current_cluster <- .y$Cluster
    #Print Cluster 
    print(paste("------------------------CLUSTER ", current_cluster, "--------------------------"))
    
    if(current_cluster == 2) { # For detected zero-inflation
      model <- glmmTMB(cell_per_area ~ Metadata_Pt_type * Metadata_Cortex_region * Metadata_Layer + 
                         (1|Patient/Metadata_Sample_number),
                       ziformula = ~1,   # Zero-inflation formula (constant probability of zero, no predictors)
                       family = nbinom1(link = "log"),   # Positive values modeled with negative binomial type 1
                       data = .x)
    } else {
      # For other clusters, apply the regular linear mixed model (no hurdle or zero-inflation)
      model <- glmmTMB(cell_per_area ~ Metadata_Pt_type * Metadata_Cortex_region * Metadata_Layer + 
                         (1 | Patient/Metadata_Sample_number), 
                       family = nbinom1(link = "log"), 
                       data = .x)
    }
    #Model performance
    print(model)
    windows(width = 15, height = 15)
    print(check_model(model))
    print(check_overdispersion(model))
    print(check_zeroinflation(model))
    
    # Perform emmeans for later extraction
    emm <- emmip(model, Metadata_Pt_type ~ Metadata_Layer + Metadata_Cortex_region, plotit = FALSE, CIs = TRUE, type="response")
    
    #Evaluate Contrasts
    #Metadata_Pt_type
    emm_Metadata_Pt_type <- emmeans(model, pairwise ~ Metadata_Pt_type | Metadata_Layer + Metadata_Cortex_region)
    print(emm_Metadata_Pt_type$contrasts)
    #Metadata_Layer
    emm_Metadata_Layer <- emmeans(model, pairwise ~ Metadata_Layer | Metadata_Pt_type + Metadata_Cortex_region)
    print(emm_Metadata_Layer$contrasts)
    #Metadata_Cortex_region
    emm_Metadata_Cortex_region <- emmeans(model, pairwise ~ Metadata_Cortex_region | Metadata_Layer + Metadata_Pt_type)
    print(emm_Metadata_Cortex_region$contrasts)
    
    # Extract estimates, SE and CI
    estimates_se <- as.data.frame(emm)
    
    return(estimates_se)
  }) %>%
  ungroup()  # Remove grouping for final combined dataframe

```

```{r}
#Prepare data for plotting
statistics_n_cluster$Metadata_Pt_type <- factor(statistics_n_cluster$Metadata_Pt_type, levels = c("CTRL", "AD", "CJD"))
statistics_n_cluster$Metadata_Layer <- factor(statistics_n_cluster$Metadata_Layer, levels = c("sGM", "dGM", "WM"))

y_limit <- max(total_clusters_per_image$cell_per_area, na.rm = TRUE)

#Plot
p <- ggplot(statistics_n_cluster, aes(x=Cluster, y=yvar)) +
  facet_grid("Cortex Layer" + Metadata_Layer~"Brain Region" + Metadata_Cortex_region)+
  geom_bar(aes(x=Cluster, y=yvar, fill=Metadata_Pt_type, color=Cluster), stat="identity", 
           position=position_dodge(0.9), size=0.75, alpha=0.6) +
  geom_errorbar(aes(fill = Metadata_Pt_type, ymin=LCL, ymax=UCL, color = Cluster), stat="identity", width=.2,
                position=position_dodge(0.9), color = "#595959", alpha = 1)+
  theme(panel.grid.major = element_line(size = 1))+
  scale_fill_manual(values=patient_group_colors)+
  scale_color_manual(values = cluster_colors, name = "IBA1+ Object Cluster",  
                     labels = c("1", "2.1", "2.2"))+
  ylab(expression("Number of IBA1+ Object Clusters /" ~ mm^2))+
  xlab("IBA1+ Object Cluster")+
  guides(fill=guide_legend(title="Estimated Mean"))+
  theme(plot.title = element_text(hjust = 0.5))+
  expand_limits(x = 0, y = 0)+
  ggtitle("")+
  theme(text = element_text(size = 10, face="bold"),strip.text = element_text(face = "bold"),
        legend.text = element_text(size = 6), 
        legend.title = element_text(size = 8),
        legend.key.size = unit(0.5, "cm"),
        legend.position = "bottom",         
        legend.direction = "horizontal",
        legend.title.position = "top")+
  scale_x_discrete(labels = c("1" = "1", "2" = "2.1", "3" = "2.2"))+
  guides(color=guide_legend(override.aes=list(fill=NA)))
p
```



#Stacked Bar Chart of Distribution of the total object population
```{r}
total_cells_per_group <- single_profile %>% group_by(Metadata_Pt_type, Metadata_Cortex_region, Metadata_Layer) %>% count() # Total Object Count
colnames(total_cells_per_group)[4] <- "n_total_cell_count"#Rename

grouped_cluster_count <- single_profile  %>% group_by(Metadata_Pt_type, Metadata_Cortex_region, Metadata_Layer) %>%
  count(Cluster) #Cluster count per permutation group
colnames(grouped_cluster_count)[5] <- "n_cluster_in_group" #rename

cluster_distribution <- left_join(grouped_cluster_count, total_cells_per_group, 
                                  by = c("Metadata_Pt_type"="Metadata_Pt_type", "Metadata_Cortex_region"="Metadata_Cortex_region", "Metadata_Layer"="Metadata_Layer"))%>% 
  mutate(percentage = (n_cluster_in_group/n_total_cell_count)*100) #Get Percentage of Each Cluster in each permutation group

#Reorder
cluster_distribution$Metadata_Pt_type <- factor(cluster_distribution$Metadata_Pt_type, levels = c("CTRL", "AD", "CJD"))
cluster_distribution$Metadata_Layer <- factor(cluster_distribution$Metadata_Layer, levels = c("sGM", "dGM", "WM"))

#Stacked Bar Chart
p_stacked <- ggplot(cluster_distribution, aes(x=Metadata_Pt_type, fill = Cluster, y=percentage))+
  facet_grid("Cortex Layer" + Metadata_Layer ~ "Brain Region" + Metadata_Cortex_region)+
  geom_bar(position="fill", stat="identity")+
  scale_fill_manual(values = cluster_colors, labels = c("1", "2.1", "2.2"))+
  ylab("Proportion (%) of IBA1+ Object Clusters") +
  xlab("Disease Group")+
  scale_y_continuous(labels = percent)+
  guides(fill=guide_legend(title="IBA1+ Object Cluster"))+
  theme(text = element_text(size = 10, face="bold"),strip.text = element_text(face = "bold"),
        legend.text = element_text(size = 6), 
        legend.title = element_text(size = 8),
        legend.key.size = unit(0.5, "cm"),
        legend.position = "bottom",         
        legend.direction = "horizontal",
        legend.title.position = "top")
p_stacked
```


#Correlation with GeoMx Protein Data
```{r}
#Prepare Cluster data
count_per_image <- single_profile %>% group_by(ImageNumber) %>% count(Cluster) # Total Object count
count_per_image <- count_per_image %>% 
  left_join(total_cells_per_image, by = join_by(ImageNumber == ImageNumber))%>%
  mutate(clusterpercentage_per_image = (n/n_cells_in_image)*100)%>%
  select(-c("n_cells_in_image")) #Add cluster data, calculate percentage
count_per_image_merged <- merge(count_per_image, total_clusters_per_image, by=c("ImageNumber", "Cluster")) #merge with additional from previous analysis
count_per_image_merged <- reshape(count_per_image_merged, idvar = "ImageNumber", timevar = "Cluster", direction="wide", sep = "_") # To wide Format
count_per_image_merged[is.na(count_per_image_merged)] <- 0 #Replace NaNs with zero 


#Rename columns for merging
colnames(count_per_image_merged)[8] <- "ROI (Label)"
colnames(geomx_data)[6] <- "ROI (Label)"

#Combine GeoMx and Cellprofiler Distribution data
combined_data <- merge(geomx_data, count_per_image_merged, by = "ROI (Label)")

#Rename Column names
combined_data <- combined_data %>% rename("IBA1+ Object Cluster 1 Proportion" = clusterpercentage_per_image_1,
                                          "IBA1+ Object Cluster 2.1 Proportion" = clusterpercentage_per_image_2,
                                          "IBA1+ Object Cluster 2.2 Proportion" = clusterpercentage_per_image_3)

#Isolate columns
columns <- c("IBA1+ Object Cluster 2.2 Proportion", "IBA1+ Object Cluster 2.1 Proportion", "IBA1+ Object Cluster 1 Proportion",
             "HLA DR", "CD11c", "CD9", "CD68", "CD39", "CD40", "Ki 67", 
             "IBA1", "CD45", "CD11b", "P2ry12", "TMEM119", "MERTK", "CD31", "CSF1R", "CTSD", "GPNMB")

#Repeated measurement correlation
dist_rmc_mat <- rmcorr_mat(participant = Patient, 
                           variables = columns,
                           dataset = combined_data,
                           CI.level = 0.95)

#Isolate correct correlations
isolated_matrix <- dist_rmc_mat$matrix[-c(1:3), c(1:3)]

#Visualization
ggcorrplot(isolated_matrix, method = "circle") + 
  theme(legend.title = element_text(size = 10),legend.text = element_text(size = 8))+
  guides(
    fill = guide_colorbar(
      title = expression(bold(r)[bold(rm)]),                  
      barwidth = 1,                      
      barheight = 5                     
    ))
```


#Save tab-delimited data for MoBIE
```{r}
cluster_ID <- single_profile %>% select(ImageNumber, ObjectNumber, Cluster)
write.delim(cluster_ID, here("Data", "Cellprofiler (morphology)","MoBIE", "Microglia_clusterID.txt"), 
            row.names=FALSE, sep = "\t")
```

